Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-10-14 ms/plos_latex.tex
working ...

\end{flushleft}
% Please keep the abstract below 300 words
% current is 298!
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
  course of an emerging epidemic have important implications both for
  our basic understanding of epidemiological dynamics and,
  potentially, for the outcomes of public health interventions.
  Changes in the fitness landscape
  will generally select for higher virulence during the early
  phase of an epidemic, but quantitative outcomes
  can depend sensitively on biological details and the structure of
  mathematical models used to capture them.  Fraser, Shirreff, and
  co-workers have proposed a series of models for eco-evolutionary
  dynamics of HIV that are relatively detailed in their portrayal of
  the tradeoffs between transmission and virulence (mediated by
  set-point viral load, SPVL) and their heritability between
  hosts. However, these models use implicit
  representations of the transmission process that drastically simplify the
  partnership dynamics that previous research has found to be critical
  in driving epidemics of sexually transmitted diseases.  We
  explore models that combine HIV virulence tradeoffs with a range of
  epidemiological structures, modeling partnership formation and
  dissolution and allowing for individuals to transmit disease outside
  of partnerships. We assess summary statistics such as the peak value
  of virulence (SPVL) and the time at which the peak occurs across all
  models for a range of 
  partnership dynamic parameters 
  applicable to the HIV epidemic in sub-Saharan Africa. In
  order to account for the different interpretations of parameters
  across model structures, we scale all parameter sets to constrain
  the simulated epidemic growth rate to be identical, matching a
  realistic baseline value. 
  Although SPVL trajectories are broadly similar
  across model structures --- an initial increase in SPVL, 
  followed by a peak and a decline to an equilibrium SPVL slightly
  lower than a peak --- the timing and height of the peak vary
  considerably.
  Models of intermediate
  complexity predicted much lower peak virulence (15 years to
  progress to AIDS) compared to both more realistic models
  and simple random-mixing models with no partnership structure
  at all (both approx. 7.25 years to progress to AIDS).
  In this range of models, the simplest random-mixing structure is
  actually the best approximation to the most realistic model; this
  surprising outcome occurs because the dominance of extra-pair
  contact in the realistic model tends to mask the effects of
  partnership structure.

@@

The evolution of pathogen virulence is a fundamental process in
evolutionary biology, of both theoretical and (potentially) practical
importance. In general, evolutionary suggests that
disease strains that can reproduce more, where reproduction
means the amount of \emph{between-host} transmission, or
the number of new hosts infected, will increase in prevalence.
Pathogens can increase their between-host 
fitness, or net reproduction rate, either
by increasing their birth rate, the rate at which they
infect new hosts, or by decreasing their death rate, the rate
at which hosts recover or die from disease.
The trade-off theory \cite{Ebert1999} --- postulates that
the birth and death rate are both linked to the rate at which
the pathogen reproduces the host, and that the pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
The tradeoff theory has been criticized
\cite{EbertBull2003,alizon_adaptive_2015}, but has also been
successfully applied in a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}. One
particularly interesting application of these ideas is the work by
Fraser \etal\ showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in studies of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
transmission
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

@@

We add complexity to the epidemiological model following the general approach of Champredon \etal\ \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships. In order to explore how virulence evolution depends on epidemiological structure, we consider a series of models with increasing levels of complexity. In order to avoid dependence of the results on a particular set of parameters --- as we explain below, finding matching sets of parameters across models with widely differing epidemiological structures is challenging --- we evaluate our models across a wide range of parameters, again following Champredon \etal\ \cite{champredon_hiv_2013} in using a Latin hypercube design. For each model run, we compute a set of metrics (peak virulence, timing of virulence peak, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.

@@

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\end{equation}
where $V_{\alpha} = 10^\alpha$. 
The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix). Although such simplification affects evolution of virulence, we show that our simplified model is able to produce result that is qualitatively similar to that of Shirreff \etal's \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}). As our focus is not making accurate predictions, we believe that being able to produce qualitatively similar dynamics is sufficient for this study.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
% current is 298!
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
  course of an emerging epidemic have important implications both for
  our basic understanding of epidemiological dynamics and,
  potentially, for the outcomes of public health interventions.
  Changes in the fitness landscape
  will generally select for higher virulence during the early
  phase of an epidemic, but quantitative outcomes
  may depend sensitively on biological details and the structure of
  mathematical models used to capture them.  Fraser and
  co-workers have proposed a series of models for eco-evolutionary
  dynamics of HIV that are relatively detailed in their portrayal of
  the tradeoffs between transmission and virulence (mediated by
  set-point viral load, SPVL) and their heritability between
  hosts. However, these models use implicit
  representations of the transmission process that drastically simplify the
  partnership dynamics that previous research has found to be critical
  in driving epidemics of sexually transmitted diseases.  We
  explore models that combine HIV virulence tradeoffs with a range of
  epidemiological structures, modeling partnership formation and
  dissolution and allowing for individuals to transmit disease outside
  of partnerships. We assess summary statistics such as the minimum
  time of progression to AIDS and the time at which this 
  extreme value (peak virulence) occurs across all
  models for a range of 
  partnership dynamic parameters 
  applicable to the HIV epidemic in sub-Saharan Africa. In
  order to account for the different interpretations of parameters
  across model structures, we scale all parameter sets to constrain
  the simulated epidemic growth rate to be identical, matching a
  realistic baseline value. 
  Although virulence trajectories are broadly similar
  across model structures --- an initial increase in virulence, 
  followed by a peak and a decline to an equilibrium virulence slightly
  lower than a peak --- the timing and magnitude of the peak vary
  considerably.
  Models of intermediate
  complexity predicted much lower peak virulence (15 years to
  progress to AIDS) compared to both more realistic models
  and simple random-mixing models with no partnership structure
  at all (both approx. 7.25 years to progress to AIDS).
  In this range of models, the simplest random-mixing structure is
  actually the best approximation to the most realistic model; this
  surprising outcome occurs because the dominance of extra-pair
  contact in the realistic model tends to mask the effects of
  partnership structure.

@@

The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
disease strains that can reproduce more --- where reproduction
is defined the amount of \emph{between-host} transmission, or
the number of new hosts infected --- will increase in prevalence.
Pathogens can increase their net reproduction rate either
by increasing their birth rate, the rate (per infected host) at which they
infect new hosts, or by decreasing their death rate, the rate
at which hosts recover or die from disease.
The trade-off theory \cite{anderson_coevolution_1982,Ebert1999} --- postulates that
the birth and death rate are both linked to the rate at which
the pathogen reproduces within the host, and that the pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
Fraser \etal\ have applied these ideas in a particularly
interesting way 
by showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in studies of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
transmission
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

@@

We add complexity to the epidemiological model following the general
approach of Champredon \etal\ \cite{champredon_hiv_2013},
which is in turn based on work of Dietz and Hadeler
\cite{dietz_epidemiological_1988}; individuals
join and leave partnerships at a specified rate, and can have sexual
contact both within and outside of established partnerships. In order
to explore how virulence evolution depends on epidemiological
structure, we consider a series of models with increasing levels of
complexity. In order to avoid dependence of the results on a
particular set of parameters --- as we explain below, finding matching
sets of parameters across models with widely differing epidemiological
structures is challenging --- we evaluate our models across a wide
range of parameters, again following Champredon
\etal\ \cite{champredon_hiv_2013} in using a Latin hypercube
design. For each model run, we compute a set of metrics (peak
virulence, timing of virulence peak, equilibrium virulence) that
summarize the evolutionary trajectory of a simulated HIV epidemic.

@@

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\end{equation}
where $V_{\alpha} = 10^\alpha$. 
The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix). Although such simplification could certainly affect the evolution of virulence (Kret\cite{kretzschmar_effect_1998} , our simplified model is able to produce qualitatively similar results that is qualitatively similar to that of Shirreff \etal's \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}). As our focus is not making accurate predictions, we believe that being able to produce qualitatively similar dynamics is sufficient for this study.
